## Review of Personal Monitoring Device

Due to the ever increasing capacity to collect large amounts of data, personal data devices such as Fitbit and Jawbone are increasing becoming a part of everyday life.  These particular devices take measurments in the quest to find behavior patterns and to improve ones health.  

The review uses data from a personal activity device that collected data at 5 minute intervals, during the month of October and November 2012. 
<!-- For more info on RMarkdown see http://rmarkdown.rstudio.com/ -->

### Data

The data variables include:

* steps - Number of steps taken in 5 minute intervals.
* date - A date the measurements were taken.
* interval - 5 minute interval identifier.

### Data Processing
  
```{r libraryload, echo=TRUE }
library("data.table") # Data storage.
library("assertthat") # Useful test.
library("chron")      # Convenient date storage.
library("Hmisc")      # Check attributes.
library("mice")       # mice imputation. 
```
```{r, personaloptions}
# List of options.
Sys.setlocale()
op <- options()
options(digits = 3)    
```
The analysis begins by defining a monitor object which contain member variables and functions that are members of the monitor class.
```{r monitorclass}
# Monitor class definition and functions.
getName <- function(.Object) {return(.Object@activityName)}
getActivity <- function(.Object) {return(.Object@activity)}
getClass <- function(.Object) {return(sapply(.Object@activity, class))}
summary <- function(.Object) {
  # Calculates mean, median, and total number of steps each day.
  # Total steps per day with NAs and zeros removed.
  # Zeros removed to facilitate median calculations.
  .Object@activity[!is.na(steps) & (steps > 0), 
                                   .(count = .N,
                                   steps = sum(steps),
                                   min = min(steps),
                                   max = max(steps),
                                   mean = mean(steps),
                                   median = median(steps)),
                                   by = date]
}
# totalMissing returns number of NAs. 
totalMissing <- function(.Object) {return(sum(is.na(.Object@activity)))}

setClass(
  Class = "Monitor", 
  slots = c(
    name = "character", 
    activity = "data.table"
  )
)
setGeneric(
  name = "gname", 
  def = function(.Object) {standardGeneric("gname")}
)
setGeneric(
  name = "gactivity",
  def = function(.Object) {standardGeneric("gactivity")}
) 
setGeneric(
  name = "gclass",
  def = function(.Object) {standardGeneric("gclass")}
)
setGeneric(
  name = "gsummary",
  def = function(.Object) {standardGeneric("gsummary")}
)
setGeneric(
  name = "totMissing",
  def = function(.Object) {stanardGeneric("totMissing")}
  )
setMethod("gname",signature("Monitor"), getName)
setMethod("gactivity",signature("Monitor"), getActivity)
setMethod("gclass",signature("Monitor"), getClass)
setMethod("initialize", "Monitor", function(.Object, ...) {
  .Object <- callNextMethod()
  # Automatically sets classes on object construction. 
  # as.chron reads dates as numeric.                                           
  .Object@activity[,date := as.chron(as.IDate(date))] 
  .Object@activity[,steps := as.numeric(steps)]
  .Object@activity[,interval := as.numeric(interval)]
  # Create weekday and weekend variables. 
  .Object@activity[,wkdy := weekdays(date, abbreviate = TRUE)]
  .Object@activity[,day := {ifelse(weekdays(date) == "Sat","weekend",
          ifelse(weekdays(date) == "Sun", "weekend", "weekday"))},]
  # Factor variables.
  .Object@activity[,day := factor(day)]
  .Object@activity[, wkdy := factor(wkdy)]
  .Object
})
setMethod("gsummary",signature("Monitor"), summary)
setMethod("totMissing",signature("Monitor"), totalMissing)
```

```{r helperfunctions}
createDir <- function(dir_name, file_name) {
  # Create directory to hold downloaded file.
  if(!file.exists(dir_name)) {
    dir.create(dir_name)
  }
  # Unzip file into data directory.
  if(file.exists(file_name) & has_extension(file_name, "zip")) {
    unzip(zipfile = file_name, exdir = path.expand(dir_name))
    list.files(dir_name)
    return(TRUE)
  }
}
# Read dataset function.
loadDataset <- function(input, ...) {
  dataset = NULL
  try(dataset <- fread(input = input, ...), silent = TRUE)
  cat(sprintf("Processing dataset...\n"))
  return(dataset)
}
# Read data into monitor class.
monitorDv <- function(activity) {
  if(missing(activity)) {
    createDir("data", "repdata-data-activity.zip")  
    data <- loadDataset("./data/activity.csv", header = TRUE, sep = ",")
    new(Class = "Monitor", activity = data)  
  }
  else {
    new(Class = "Monitor", activity = activity)
  }
}
```

### Load data
```{r constructor, cache = TRUE}
# Construct a personal activity monitor device.
pamd <- monitorDv()
# Return activity data.
activity <- getActivity(pamd)
```

### Check data
```{r datachecks}
getClass(pamd)
dim(activity)
names(activity)
```
### Mean total number of steps per day
```{r histogramsteps, fig.width = 4, fig.height = 4, cache = TRUE}
# Returns mean and median calculations of total number of steps per day.
(s <-summary(pamd))
# Histogram of total number of steps each day
(stp <- hist(s$steps, breaks = "Sturges", plot = FALSE))
plot(stp, border = "black", col = "limegreen", 
     main = "Histogram of total steps",
     xlab = "Number of Steps", labels = TRUE,
     ylim=c(0, 30))
     # Distribution of values, jittered. 
     rug(jitter(s$steps, amount = 0.01))
```

### Average daily pattern
```{r timeseriesplot, fig.width = 5, fig.height = 4, cache = TRUE}
# Time series plot of 5 minute interval averaged over number of steps.
a <- activity[!is.na(steps), .(steps = mean(steps)), by = interval]
par(lty = 1, pch = 19)
plot(a$steps ~ a$interval, 
     type = "l", lwd = 1.3, col = "darkgreen",
     main = "Average daily steps activity",
     sub = "5 minute intervals",
     xlab = "Interval", ylab = "Number of Steps (averaged)", bg = "seashell")
```
```{r maxinterval}
# 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.
a[which.max(a$steps),]
```


### Imputation
```{r imputation, cache = TRUE}
# Return number of rows that contain NAs. 
totalMissing(pamd)
sum(activity[,!complete.cases(activity)])
library("VIM")
aggr(activity, prop = FALSE, numbers = TRUE)

# Imputation
# Another way to count missing values.
md.pattern(activity)

# Multiple imputation with chain equations to calculate missing values.
# One method is to use predictive mean matching (pmm), which is the default for numerical values for mice. 
# Dry run in order to change default values.
ini <- mice(activity, maxit = 0)
pred <- ini$predictorMatrix
# Remove weekend and weekday variables as predictors.
pred[, "day"] <- 0
pred[, "wkdy"] <- 0
# Create 3 multiple imputations and store datasets in imp object.
imp <- mice(activity, pred = pred, pri = FALSE, m = 3, seed = 4321)
```

```{r stepsimputation, eval = TRUE}
# Steps imputation stored here. 
# Each column represents the three imputted dataset.
# Show the first ten rows.
imp$imp$steps[1:10,]
```

```{r diagnosis, cache = TRUE}
# A scatterplot and stripplot are drawn that combines steps and interval for each imputted dataset.
# Red points represent the imputted values while blue points represent the observed values.
# The red points practically overlap the blue points leading to possible measurements.  
# However, additional tests should be carried out to validate measurements.
stripplot(imp, pch = 20, cex = 1.2)
```
```{r diagnosis2, eval = TRUE, cache = TRUE}
xyplot(imp, steps ~ interval | .imp, pch = 20, cex = 1.4)
```
```{r completedataset}
# A new dataset with missing data filled in is created.
# The complete dataset includes inputted data with observed data.
# The first dataset is selected.
comDt <- as.data.table(complete(imp, 1))
```
```{r meanimputted }
# Mean and median calculation using imputted data.
(s1 <- comDt[(steps > 0), 
            .(
            mean = mean(steps),
            median = median(steps),
            steps = sum(steps)
            ), 
            by = date])
```
```{r histimp, fig.width = 4, fig.height = 4, cache = TRUE}
# Histogram of imputted data.
(stp <- hist(s1$steps, breaks = "Sturges", plot = FALSE))
plot(stp, border = "black", col = "limegreen", 
     main = "Histogram of total steps", 
     xlab = "Number of Steps", labels = TRUE,
     ylim=c(0, 35))
     rug(jitter(s1$steps, amount = 0.01))
# Values in the range of 5000 to 15000 steps have increased. 
# One result of imputting missing data is the ability to use more data thus leading to a 
# more complete assessment.
```

### Activity patterns on weekend and weekdays

```{r weekendpatterns, cache = TRUE}
# Dataset with imputted variables.
s2 <- comDt[, 
     .(
     steps = sum(steps)
     ), 
     by = .(interval, day)]
```
```{r weekendplot}
xyplot(steps ~ interval | day,
    data = s2, 
    layout = c(1,2),
    panel = function(x, y) {
    panel.grid(h = -1, v = 2)
    panel.xyplot(x, y, type = "l")
})
```

